Review Paper Draft

### 2020-06-23 main.tex
Section 3 small tweaks - in progress

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects belonging to $q \le p$ subpopulations and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$ \pb{do we need to explicitly assume q discrete subpopulations here?} \anna{In order to show the equivalence of $u \sim N(0, \sigmagg \bK) \text{ and } Z\gamma \sim N(0, \sigmagg \bK)$ I think so, but maybe I can point out that this is the assumption the equivalence is arising from, but we don't' have to have discrete suppopulations in practice...}. We denote the standardized version \pb{need more explicit details on standardization} \anna{get details from Hayes 2009} of this genotype matrix as $\bX$. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bZ$ is an $n \times q$ matrix that indicates population structure, $\bgamma$ is an $q \times 1$ vector of subpopulation-specific environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$.

Both PC-Lasso and LMM-Lasso rely on matrix decompositions of $\bX$ and realized relationship matrix (RRM), $\bK$, which describes the pair-wise genetic similarity among subjects \cite{hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM:
\begin{align}
    \label{eqn:k}
    \bK &= \frac{1}{p} \bX \bX^T \nonumber \\
                   &= \frac{1}{p} \bU \bD^2 \bU^T \nonumber \\
                   &= \frac{1}{p} \bU \bD (\bU \bD)^T \nonumber \\
                   &\propto \bR \bR^T,
\end{align}
where the columns of $\bU$ are the principal components of $\bX$ (equivalently, those of $\bK$) and $\boldsymbol{D}^2$ is a diagonal matrix of the eigenvalues of $\bX$ \pb{only square matrices have eigenvalues} (equivalently, the singular values of $\bK$.) The columns of $\boldsymbol{R}$ are therefore proportional to the principal components weighted by their corresponding singular values.

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \cite{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \ref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\boldsymbol{\beta, \alpha| X, y}) = \frac{1}{2n} ||\boldsymbol{y} - \boldsymbol{X \beta} - \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \boldsymbol{\beta} ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients. \anna{do I need to define the ell2 norm?} \pb{YES} Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects requires the estimation of $p + k + 1$ parameters.

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming $\bgamma \sim N(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p)$ \pb{too many sub/superscripts for an in-line fraction}. Typically, $\bZ$ is not observed directly, and a common convention \citep[e.g.,][]{Rakitsch2012} is to replace $\bZ$ with $\bX$. Then, rather than assuming anything about $\bgamma$ directly, let $\bu=\bX\bgamma$ represent the random effect and assume $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigmagg \bK)$. This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$ and thus $\by  \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso then performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}}\bX \bbeta, \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
Using the spectral decomposition of $\bK$ described in equation \ref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization \pb{ONLY WORKS IF $\bX$ IS FULL RANK}
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$, $\byT = \bU^T \by$, and $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \cite{Rakitsch2012} uses a two-step procedure for estimating the variance components, in which the variance components and corresponding weights are estimated once under the assumption of a null model \pb{I think you should be more specific here about the ``variance components'', use the $\sigma_e^2$ notation you just introduced, etc.}. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \cite{glmnet} or \texttt{ncvreg} \cite{ncvreg}. LMM-Lasso-ggmix \cite{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \cite{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to the results section.

>>>>

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects belonging to $q \le p$ subpopulations and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$ \pb{do we need to explicitly assume q discrete subpopulations here?} \anna{In order to show the equivalence of $u \sim N(0, \sigmagg \bK) \text{ and } Z\gamma \sim N(0, \sigmagg \bK)$ I think so, but maybe I can point out that this is the assumption the equivalence is arising from, but we don't' have to have discrete suppopulations in practice...\textbf{do this}}. We denote the standardized version \pb{need more explicit details on standardization} \anna{\textbf{do this} get details from Hayes 2009} of this genotype matrix as $\bX$. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,
\begin{equation}
    \label{eqn:model}
    \by = \bX\bbeta + \bZ\bgamma + \beps
\end{equation}
where $\bbeta$ is a $p \times 1$ vector of SNP effects, $\bZ$ is an $n \times q$ matrix that indicates population structure, $\bgamma$ is an $q \times 1$ vector of subpopulation-specific environmental effects, and $\beps \sim N(\boldsymbol{0}, \sigmaee \bI_n)$.

Both PC-Lasso and LMM-Lasso rely on matrix decompositions of $\bX$ and realized relationship matrix (RRM), $\bK$, which describes the pair-wise genetic similarity among subjects \cite{hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM: \anna{does this sentence need to change?}
\begin{align}
    \label{eqn:k}
    \bK &= \frac{1}{p} \bX \bX^T \nonumber \\
                   &= \frac{1}{p} \bU \bD^2 \bU^T \nonumber \\
                   &= \frac{1}{p} \bU \bD (\bU \bD)^T \nonumber \\
                   &\propto \bR \bR^T,
\end{align}
where the columns of $\bU$ are the principal components of $\bX$ (equivalently, those of $\bK$) and $\boldsymbol{D}^2$ is a diagonal matrix of the singular values of $\bK$. The columns of $\boldsymbol{R}$ are therefore proportional to the principal components weighted by their corresponding singular values.

%------------------------------%
\subsection{PC-Lasso}
PC-Lasso uses the first $k$ principal components of $\bX$ to approximate $\bZ\bgamma$ and includes them in the model as unpenalized covariates. It has been shown \cite{hoffman2013correcting} that this corresponds to the regression model
\begin{equation}
    \label{eqn:pca_reg}
    \by = \bX\bbeta + \bU_{1:k}\boldsymbol{\alpha} + \beps 
\end{equation}
where $\bU_{1:k}$ is the $n \times k$ matrix of principal component vectors, $\boldsymbol{\alpha}$ their corresponding $k \times 1$ coefficient vector, and all other parameters are as defined in equation \ref{eqn:model}. In a penalized regression framework, this model yields the objective function
\begin{equation}
    \label{eqn:pca_obj}
    Q(\boldsymbol{\beta, \alpha| X, y}) = \frac{1}{2n} ||\boldsymbol{y} - \boldsymbol{X \beta} - \bU_{1:k}\boldsymbol{\alpha}||_2^2 + \lambda || \boldsymbol{\beta} ||_1
\end{equation}
where $\lambda$ is a tuning parameter, and $|| \boldsymbol{\beta} ||_1 = \sum_{j=1}^p |\beta_j|$ denotes the $\ell_1$ norm of the regression coefficients, \anna{and $||.||_2$ the $\ell_2$ norm.} Note that the principal component regression coefficients, $\boldsymbol{\alpha}$, are not included in the penalty term. Modeling unobserved confounding using fixed effects in this manner thus requires the estimation of $p + k + 1$ parameters.

%------------------------------%
\subsection{LMM-Lasso}
LMM-Lasso models $\bZ\bgamma$ as a random effect assuming 
$$\bgamma \sim N(\boldsymbol{0}, \frac{\sigmagg}{p} \bI_p).$$  Typically, $\bZ$ is not observed directly, and a common convention \citep[e.g.,][]{Rakitsch2012} is to replace $\bZ$ with $\bX$. Then, rather than assuming anything about $\bgamma$ directly, let $\bu=\bX\bgamma$ represent the random effect and assume $\boldsymbol{u} \sim N(\boldsymbol{0}, \sigmagg \bK)$. This results in the LMM regression model
\begin{equation}
    \label{eqn:lmm_reg}
    \by = \bX\bbeta + \bX\bgamma+ \beps,
\end{equation}
where we assume $\beps \independent \bX\bgamma$ and thus $\by  \sim N(\bX \bbeta, \bV)$, where $\bV = \sigmagg \textbf{K} + \sigmaee \textbf{I}$. LMM-Lasso then performs a weighted rotation of the original data, left-multiplying by $\bV^{-1/2}$ in a manner akin to generalized least squares to diagonalize the error term such that
\begin{equation}
\bV^{-\frac{1}{2}}\by \sim N(\bV^{-\frac{1}{2}}\bX \bbeta, \bI).
\end{equation}
\anna{...however, as Karl Rohe paper shows, GLS does not completely translate to penalized regression...comment on this?}
Using the spectral decomposition of $\bK$ described in equation \ref{eqn:k}, we can gain further insight into the mechanism of this transformation. Note that $\bV^{-1}$ can be expressed using the following matrix factorization \pb{ONLY WORKS IF $\bX$ IS FULL RANK} \anna{I understand this on paper, but why does it work numerically? Even if $n > p$, svd(K) or eigen(K) gives a U that is $n \times n$... what are the vectors corresponding to the 0-eigenvalues? Toy example...}
\begin{align*}
    \bV^{-1} &= (\sigmagg \bK + \sigmaee \bI)^{-1}\\
    &=(\sigmagg \bU \bD^2 \bU^T + \sigmaee \bU \bU^T)^{-1}\\
    &= (\bU (\sigmagg \bD^2 + \sigmaee \bI) \bU^T)^{-1}\\
    &=\bU (\sigmagg \bD^2 + \sigmaee \bI)^{-1} \bU^T.
\end{align*}
By incorporating these decompositions, the transformed data can be expressed as $\bW \bXT$ and $\bW \byT$, respectively, where $\bXT = \bU^T \bX$, $\byT = \bU^T \by$, and $\bW = \text{diag}\{w_1, ... w_n\}$ where $w_i$ corresponds to the square root of the $i$th element of the diagonal matrix $(\sigmagg \boldsymbol{D}^2 + \sigmaee \bI)^{-1}$. Together, this yields the objective function
\begin{equation}
\label{eqn:lmm_obj}
Q(\bbeta | \bXT, \byT, \bW) = || \bW (\byT - \bXT \bbeta)||_2^2 + \lambda || \bbeta ||_1.
\end{equation}
An advantage to modeling unobserved confounding using a random effect, as opposed to fixed effects as in PC-Lasso, is that this requires the estimation of substantially fewer parameters: $p + 2$ for LMM-Lasso-Rakitsch, and $p + 3$ for LMM-Lasso-ggmix.

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \cite{Rakitsch2012} uses \anna{a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$. These variance components and the corresponding weights are estimated once under the assumption of a null model.} A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \cite{glmnet} or \texttt{ncvreg} \cite{ncvreg}. LMM-Lasso-ggmix \cite{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \cite{ggmix, bhatnagar2019simultaneous}. We postpone further discussion of the differences between these methods to the results section.
